---
title: "multi_var_analy"
author: "Lxy"
date: "November 11, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Factor analysis

[^1]: 
* Johnson RA & Wichern DW, Applied Multivariate Statistical Analysis, 6ed, Pearson,. Chapter 9
* Rencher AC & Christensen WF, Methods of Multivariate Analysis, 3ed, Wiley. Ch 13.

Consider random vector $\underline{X}^{T} = (X_{1}, ... , X_{p}) \in \mathbb{Z}$ where $E(\underline{X}) = \underline{\mu}_{p \times 1}$ and $Var{\underline{X}} = \Sigma_{p \times p}$

$$ X_{j} - \mu_{j} = \lambda_{j1}F_{1} + ... + \lambda_{jm}F_{m} + \epsilon_{j}, j = 1, ..., p $$

*F* are common factors and $\epsilon$ are specific factors. Both are random.

Factor model in matrix form:
$$ \mathbf{\underline{X} - \underline{\mu} = \Lambda\underline{F} + \underline{U}} $$
$$ \underline{\mu} = mean\ vector $$
$$ \Lambda = (p \times p)\ loadings $$
$$ \underline{F} = (m \times 1)\ common\ factors,\ E(\underline{F}) = \underline{0},\ Var(\underline{F}) = I_{m} $$
$$ \underline{U} = (p \times 1)\ specific\ factors,\ E(\underline{U}) = \underline{0},\ Var(\underline{U}) = \Psi = diag(\psi_{1}, \dots , \psi_{p}) $$
$$ Cov(\underline{F}, \underline{U}) = 0_{m \times p} $$

Then we have $Var(\underline{X}) = \Sigma = (\sigma_{jj})_{p \times p}= \Lambda\Lambda^{T} + \Psi$ which is called the **factor analysis equations**.

$$ \sigma_{jj} = \sum_{k=1}^m \lambda_{jk}^{2} + \psi_{j} = h_{j}^{2} + \psi_{j} $$
$$ h_{j}^{2} is called the j-th communality and \psi_{j} is the specific variance of uniqueness $$





R does not have factor loading function utilising principal component, hence we introduce the function below.

```{r function: factor loading with principal component method}

# https://stat.ethz.ch/pipermail/r-help/2002-August/023929.html 
factanalpc <- function(xmat, factors=NULL, cor=TRUE) {
  prc <- princomp ( covmat = xmat ,cor = cor )
  eig <- prc$sdev^2
  if (is.null(factors)) factors <- sum ( eig >= 1 ) 
  loadings <- prc$loadings [ , 1:factors ]
  coefficients <- loadings [ , 1:factors ] %*% diag ( prc$sdev[1:factors] )
#   rotated <- varimax ( coefficients ) $ loadings # execute this line if rotation is wanted
  rotated <-  coefficients # output the loadings without rotation
  fct.ss  <- apply( rotated, 2 , function (x) sum (x^2) )
  pct.ss  <- fct.ss / sum (eig)
  cum.ss  <- cumsum ( pct.ss )
  ss <- t ( cbind ( fct.ss , pct.ss, cum.ss ) )
  colnames(rotated) <-c("Factor 1","Factor 2")
  colnames(ss) <- colnames(rotated)
  rownames(ss) <- c("Var explained", "Prop of var expl", "Cum percent of var expl")
  out <- list(rotated,ss)
  return (out)
}

```








